Unshuffling Permutations

### 2015-06-30 article.tex
Partie algebre (suite)

\DeclareMathOperator{\LEFT}{\ell}
\DeclareMathOperator{\RIGHT}{r}
\DeclareMathOperator{\QQ}{\mathbb{Q}}
\DeclareMathOperator{\STD}{\mathrm{std}}
\DeclareMathOperator{\SHUFFLE}{\bullet}

@@

% Définir :
% - notation $[n]$
% - ensemble $S$
% - standardisation $\STD$
% - notation $\pi_{|P}$ et $\pi_{|\bar P}$
% - notation $\pi^\sim$ (miroir)
% - notation $\bar \pi$ (complémentaire)
% - notation $\bar ^{-1}$ (inverse)

@@

\begin{proposition} \label{prop:square_stability}
    Let $\pi$ be a square permutation and $\rho$ a square root of $\pi$.
    Then,
    \begin{enumerate}[label={\it (\roman*)},fullwidth]
        \item \label{item:square_stability_1}
        the permutation $\pi^\sim$ is a square and $\rho^\sim$
        is one of its square roots;
        \item \label{item:square_stability_2}
        the permutation $\bar \pi$ is a square and $\bar \rho$
        is one of its square roots;
        \item \label{item:square_stability_3}
        the permutation $\pi^{-1}$ is a square and $\rho^{-1}$
        is one of its square roots.
    \end{enumerate}
\end{proposition}
\begin{proof}
    First of all, since $\pi$ is a square permutation and $\rho$ is a
    square root of $\pi$, $\rho \otimes \rho$ has a nonzero coefficient
    in the expansion of $\Delta(\pi)$. This implies that there are two
    disjoint sets of indexes $P_1$ and $P_2$ such that
    $P_1 \uplus P_2 = [|\pi|]$ and, by denoting $\pi_{|P_1}$ by $u$ and
    $\pi_{|P_2}$ by $v$,
    \begin{equation} \label{equ:square_stability_same_std}
        \STD(u) = \rho = \STD(v).
    \end{equation}
    \smallskip

    To prove~\ref{item:square_stability_1}, set
    $P_1^\sim := \{|\pi| - i  + 1 : i \in P_1\}$,
    $P_2^\sim := \{|\pi| - i  + 1 : i \in P_2\}$,
    $u' := \pi^\sim_{|P_1^\sim}$, and $v' := \pi^\sim_{|P_2^\sim}$.
    Now, by construction, ${u'}^\sim = u$ and ${v'}^\sim = v$. Then,
    by~\eqref{equ:square_stability_same_std},
    $\STD(u') = \rho^\sim = \STD(v')$, so that $\rho^\sim$ is a square
    root of $\pi$.
    \smallskip

@@

    TO CONTINUE

\end{proof}
\medskip

>>>>

\DeclareMathOperator{\LEFT}{\ell}
\DeclareMathOperator{\RIGHT}{r}
\DeclareMathOperator{\QQ}{\mathbb{Q}}
\DeclareMathOperator{\STD}{\mathrm{s}}
\DeclareMathOperator{\SHUFFLE}{\bullet}

@@

% Définir :
% - notation $[n]$
% - ensemble $S$
% - standardisation $\STD$
% - notation $\pi_{|P}$ et $\pi_{|\bar P}$
% - notation $\widetitle{\pi} (miroir)
% - notation $\bar \pi$ (complémentaire)
% - notation $\bar ^{-1}$ (inverse)

@@

Proposition~\ref{prop:shuffle_associative_commutative} implies that
$\QQ[S]$ endowed with the unshuffling coproduct $\Delta$ is a
coassociative cocommutative coalgebra, or in an equivalent way, that
$\QQ[S]$ endowed with the shuffle product $\SHUFFLE$ is an associative
commutative algebra.
\medskip

\begin{proposition} \label{prop:endomorphisms}
    The three linear maps
    \begin{equation}
        \phi_{\widetilde{\,}}, \: \phi_{\bar \,}, \: \phi_{-1}
        : \QQ[S] \to \QQ[S]
    \end{equation}
    linearly sending a permutation $\pi$ to, respectively,
    $\widetilde{\pi}$, $\bar \pi$, or $\pi^{-1}$ are endomorphims of associative
    algebras.
\end{proposition}
\begin{proof}
    ADAPT NEXT ARGUMENTS AND FINISH.

    First of all, since $\pi$ is a square permutation and $\rho$ is a
    square root of $\pi$, $\rho \otimes \rho$ has a nonzero coefficient
    in the expansion of $\Delta(\pi)$. This implies that there are two
    disjoint sets of indexes $P_1$ and $P_2$ such that
    $P_1 \uplus P_2 = [|\pi|]$ and, by denoting $\pi_{|P_1}$ by $u$ and
    $\pi_{|P_2}$ by $v$,
    \begin{equation} \label{equ:square_stability_same_std}
        \STD(u) = \rho = \STD(v).
    \end{equation}
    \smallskip

    To prove~\ref{item:square_stability_1}, set
    $\widetilde{P_1} := \{|\pi| - i  + 1 : i \in P_1\}$,
    $\widetilde{P_2} := \{|\pi| - i  + 1 : i \in P_2\}$,
    $u' := \widetilde{\pi}_{|\widetilde{P_1}}$, and
    $v' := \widetilde{\pi}_{|\widetilde{P_2}}$.
    Now, by construction, $\widetilde{u'} = u$ and $\widetilde{v'} = v$.
    Then, by~\eqref{equ:square_stability_same_std},
    $\STD(u') = \widetilde{\rho} = \STD(v')$, so that $\widetilde{\rho}$
    is a square root of $\pi$.
    \smallskip

@@

    TO CONTINUE
\end{proof}
\medskip

\begin{proposition} \label{prop:square_stability}
    Let $\pi$ be a square permutation and $\rho$ be a square root of
    $\pi$. Then,
    \begin{enumerate}[label={\it (\roman*)},fullwidth]
        \item \label{item:square_stability_1}
        the permutation $\widetilde{\pi}$ is a square and
        $\widetilde{\rho}$ is one of its square roots;
        \item \label{item:square_stability_2}
        the permutation $\bar \pi$ is a square and $\bar \rho$
        is one of its square roots;
        \item \label{item:square_stability_3}
        the permutation $\pi^{-1}$ is a square and $\rho^{-1}$
        is one of its square roots.
    \end{enumerate}
\end{proposition}
\begin{proof}
    Since $\pi$ is a square permutation and $\rho$ is a square root
    of $\pi$, by definition $\pi$ appears in the product
    $\rho \SHUFFLE \rho$. Now, by Proposition~\ref{prop:endomorphisms},
    since $\phi_{\widetilde{\,}}$ is a morphism of associative algebras
    from $\QQ[S]$ to $\QQ[S]$, $\phi_{\widetilde{\,}}$ commutes with
    the shuffle product of permutations $\SHUFFLE$. Hence, in particular,
    one has
    \begin{equation}
        \phi_{\widetilde{\,}}(\rho \SHUFFLE \rho) =
        \phi_{\widetilde{\,}}(\rho) \SHUFFLE \phi_{\widetilde{\,}}(\rho).
    \end{equation}
    Then, since $\pi$ appears in $\rho \SHUFFLE \rho$,
    $\widetilde{\pi}$ appears in $\widetilde{\rho} \SHUFFLE \widetilde{\rho}$.
    This implies~\ref{item:square_stability_1}.
    \smallskip

    The proofs of~\ref{item:square_stability_2}
    and~\ref{item:square_stability_3} are analogous.
\end{proof}
\medskip
